---
title: "DTP Package"
author: "[F Javier Rubio](https://sites.google.com/site/fjavierrubio67/)"
date: "25/06/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## DTP R Package

The [DTP R package](https://github.com/FJRubio67/DTP) implements the family of Double Two-Piece distributions. The family of univariate double two–piece distributions is obtained by using a density–based transformation of unimodal symmetric continuous distributions with a shape parameter. The resulting distributions contain five interpretable parameters that control the mode, as well as the scale and shape in each direction.

**References.**

1. [Bayesian modelling of skewness and kurtosis with Two-Piece Scale and shape distributions](http://projecteuclid.org/euclid.ejs/1440680330)

2. Data from: [Exploring Relationships in Body Dimensions](http://ww2.amstat.org/publications/jse/v11n2/datasets.heinz.html)

```{r}
rm(list=ls())
#library(devtools)
#install_github("FJRubio67/DTP")
library(DTP)

# Heinz et al. (2003) data. Biometric Measurements, variable "waist girth".
data <- c(66.5,61.5,61.2,78.0,70.5,66.5,58.0,74.5,73.5,90.5,66.5,61.0,69.5,62.0,63.4,71.0,59.0,74.0,60.5,75.0,65.0,82.2,70.2,85.0,
          75.0,63.0,70.0,72.6,78.7,61.0,64.2,63.1,85.0,66.0,65.0,63.2,83.6,68.0,75.7,65.5,66.9,67.9,67.8,63.2,73.1,65.3,65.6,62.7,
          81.8,72.1,68.6,80.0,64.9,61.7,64.0,58.6,68.3,69.5,70.5,69.8,69.0,81.0,71.0,73.6,96.3,65.0,62.8,83.0,67.6,65.4,61.6,63.2,
          59.4,62.7,66.6,64.6,65.7,63.0,78.5,68.2,68.5,68.8,69.0,78.0,62.0,65.8,65.0,64.9,71.5,61.4,75.4,67.2,69.5,57.9,64.4,62.3,
          68.7,68.0,66.0,71.2,64.8,78.0,60.7,75.0,68.6,67.1,68.0,71.0,76.0,70.3,69.5, 101.5,67.2,68.7,66.0,67.5,65.2,64.2,61.0,60.2,
          62.6,72.7,58.5,67.5,66.1,68.0,64.0,90.1,64.1,72.4,71.0,69.1,66.1,59.5,71.5,65.0,71.5,63.0,70.0,70.4,64.3,66.0,70.5,62.0,
          67.6,65.0,64.0,70.5,79.2,71.1,74.7,62.9,70.6,61.6,74.2,65.5,67.7,72.9,72.1,93.4,60.7,69.2,75.5,85.6,71.6,75.5,73.7,67.9,
          71.2,69.3,75.6,63.8,83.6,84.9,66.0,68.3,63.8,73.6,76.3,58.8,78.0,64.5,65.8,62.5,77.5,70.0,59.4,68.3,77.9,68.3,72.0,62.7,
          58.0,72.7,85.4,67.3,70.8,67.4,66.0,63.6,71.7,73.6,66.9,65.5,70.4,80.3,65.0,60.4,75.3,83.4,72.2,70.1,69.8,66.4,61.7,76.4,
          63.8,70.2,88.2,75.8,81.5,67.4,78.3,65.9,62.4,65.6,96.2,62.5,80.5,86.1,71.3,79.4,72.3,67.4,94.2,67.0,69.5,66.4,67.6,66.6,
          69.7,72.0,64.5,72.3,69.1,74.7,60.8,66.9,68.8,63.7,71.2,79.6,77.9,69.6,86.0,69.9,63.5,57.9,72.2,80.4)
summary(data)
n <- length(data)
#########################################################################################
# DTP-t fit
#########################################################################################
# - log-likelihood function
loglik1 <- function(par){
  if(par[2]>0 & par[3]>-1 & par[3]<1 & par[4]>0 & par[5]>0) return(- sum(ddtp(data,par[1],par[2],par[3],par[4],par[5],param="eps",dt,log=T)))
  else return(Inf)
}
# Optimisation
OPT1 <- optim(c(65,5.5,0,10,5),loglik1,control=list(maxit=10000),method="BFGS")
AIC1 <- 2*OPT1$value + 2*5
BIC1 <- 2*OPT1$value + log(n)*5
MLE1 <- OPT1$par
#########################################################################################
# TPSC-t fit
#########################################################################################
# - log-likelihood function
loglik2 <- function(par){
  if(par[2]>0 & par[3]>-1 & par[3]<1 & par[4]>0) return(- sum(ddtp(data,par[1],par[2],par[3],par[4],par[4],param="eps",dt,log=T)))
  else return(Inf)
}
# Optimisation
OPT2 <- optim(c(65,5.5,0,10),loglik2,control=list(maxit=10000),method="BFGS")
AIC2 <- 2*OPT2$value + 2*4
BIC2 <- 2*OPT2$value + log(n)*4
MLE2 <- OPT2$par
#########################################################################################
# TPSH-t fit
#########################################################################################
# - log-likelihood function
loglik3 <- function(par){
  if(par[2]>0 & par[3]>0 & par[4]>0) return(- sum(ddtp(data,par[1],par[2],0,par[3],par[4],param="eps",dt,log=T)))
  else return(Inf)
}
# Optimisation
OPT3 <- optim(c(65,5.5,10,5),loglik3,control=list(maxit=10000),method="BFGS")
AIC3 <- 2*OPT3$value + 2*5
BIC3 <- 2*OPT3$value + log(n)*5
MLE3 <- OPT3$par
#########################################################################################
# Model comparison
#########################################################################################
# AIC
c(AIC1,AIC2,AIC3)
# BIC
c(BIC1,BIC2,BIC3)
# Estimated densities
tempf1 <- function(x) ddtp(x,MLE1[1],MLE1[2],MLE1[3],MLE1[4],MLE1[5],param="eps",dt)
tempf2 <- function(x) ddtp(x,MLE2[1],MLE2[2],MLE2[3],MLE2[4],MLE2[4],param="eps",dt)
tempf3 <- function(x) ddtp(x,MLE3[1],MLE3[2],0,MLE3[3],MLE3[4],param="eps",dt)
hist(data, breaks = 20 ,probability=T,xlim=c(50,105),ylim=c(0,0.08))
curve(tempf1,50,105,add=T,lwd=2,lty=1)
curve(tempf2,50,105,add=T,lwd=2,lty=2)
curve(tempf3,50,105,add=T,lwd=2,lty=3)
box()
legend(90, 0.07, c("DTP","TPSC", "TPSH"),
       text.col = "black", lty = c(1, 2, 3),
       merge = TRUE, bg = "gray90")
```


